---
title: "Introduction to SexSeek"
description: >
  How to use SexSeek on counts matrices.
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 4.5
vignette: >
  %\VignetteIndexEntry{Introduction to SexSeek}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  dev = "png", dpi = 300, fig.retina = 1, out.width = "100%",
  dev.args = c(if (capabilities("cairo")) list(type = "cairo"), list(family = "Arial"))
)
has_tx <- requireNamespace("genevintage", quietly = TRUE) &&
  "aggregate_transcripts" %in% getNamespaceExports("genevintage")
```

```{r library}
library(SexSeek)
```

Here we use `SexSeek` on a few public GEO datasets where the sex information is available per sample from the authors
and fetched with [seqout](https://github.com/saketlab/seqout).

```{r datasets}
ExampleData()
```

## Bulk RNA-seq

### Human, gene symbols

Here we use [GSE199208](https://seqout.org/p/GSE199208), human submandibular gland, to check bulk counts where the matrix has gene symbols.

```{r human-bulk}
hb <- ExampleData("human_bulk")
dim(hb$counts)

hb_calls <- EstimateSex(hb$counts, species = hb$species, group = colnames(hb$counts))
hb_cmp <- CompareSex(hb_calls, hb$recorded)
hb_cmp[, c("unit", "verdict", "confidence", "recorded", "agrees")]
table(recorded = hb_cmp$recorded_sex, call = hb_cmp$verdict)
```

```{r human-bulk-plot}
PlotSexEvidence(hb_cmp, main = "GSE199208, human bulk")
```

### Mouse, Ensembl gene IDs

Here we use [GSE85136](https://seqout.org/p/GSE85136), mouse nucleus accumbens with both sexes in every group, to check bulk counts where the matrix has Ensembl IDs.

```{r mouse-bulk}
mb <- ExampleData("mouse_bulk")
mb_cmp <- CompareSex(
  EstimateSex(mb$counts, species = mb$species, group = colnames(mb$counts)),
  mb$recorded
)
table(recorded = mb_cmp$recorded_sex, call = mb_cmp$verdict)
```

```{r mouse-bulk-plot}
PlotSexEvidence(mb_cmp, main = "GSE85136, mouse bulk")
```

### Mouse, transcript IDs

Here we demonstrate how to use [genevintage](https://github.com/saketlab/genevintage) to first aggregate information at gene level.
Samples are taken from [GSE232600](https://seqout.org/p/GSE232600), mouse parathyroid quantified with
Salmon at the level of transcript IDs.

```{r tx-build, eval = FALSE}
map <- genevintage::transcript2gene(rownames(tx), species = "Mus musculus")
genes <- genevintage::aggregate_transcripts(tx, mapping = map, unmapped = "keep")
```


```{r tx, eval = has_tx}
mt <- ExampleData("mouse_tx")
head(rownames(mt$counts))
head(mt$tx2gene)

genes <- genevintage::aggregate_transcripts(mt$counts,
  mapping = mt$tx2gene, label = "id", unmapped = "keep", quiet = TRUE
)
dim(genes)

mt_cmp <- CompareSex(
  EstimateSex(genes, species = mt$species, group = colnames(genes)),
  mt$recorded
)
mt_cmp[, c("unit", "verdict", "confidence", "recorded", "flags")]
```

No `Xist` transcript was quantified, so females are `uncertain`:

```{r tx-notes, eval = has_tx}
unique(mt_cmp$notes[mt_cmp$verdict == "uncertain"])
```

```{r tx-plot, eval = has_tx}
PlotSexEvidence(mt_cmp, main = "GSE232600, mouse transcript level")
```

## Single-cell RNA-seq


### Human

Here we use [GSE196554](https://seqout.org/p/GSE196554), human skeletal muscle stem cells with one library per donor, to call sex per donor from single cells.

```{r human-sc}
hs <- ExampleData("human_sc")
vapply(hs$counts, ncol, 1L) # cells per donor

hs_cmp <- CompareSex(EstimateSex(hs$counts, species = hs$species), hs$recorded)
table(recorded = hs_cmp$recorded_sex, call = hs_cmp$verdict)
```

```{r human-sc-plot}
PlotSexEvidence(hs_cmp, main = "GSE196554, human single-cell (per donor)")
```

Females with `inact_frac` below 0.05 and no Y signal abstain.

```{r human-sc-uncertain}
hs_cmp[hs_cmp$verdict == "uncertain", c("unit", "inact_score", "inact_frac", "qc_score", "flags")]
```

Cells with any Y-marker count:

```{r human-sc-cells, fig.height = 4}
PlotCellFraction(hs$counts, hs$species, hs$recorded, main = "GSE196554, per donor")
```

### Mouse

Here we use [GSE132420](https://seqout.org/p/GSE132420), mouse lacrimal gland from wild-type and Sjögren's model strains, to repeat the single-cell check in mouse.

```{r mouse-sc}
ms <- ExampleData("mouse_sc")
ms_cmp <- CompareSex(EstimateSex(ms$counts, species = ms$species), ms$recorded)
table(recorded = ms_cmp$recorded_sex, call = ms_cmp$verdict)
```

```{r mouse-sc-plot}
PlotSexEvidence(ms_cmp, main = "GSE132420, mouse single-cell (per mouse)")
```

```{r mouse-sc-cells, fig.height = 4}
PlotCellFraction(ms$counts, ms$species, ms$recorded, main = "GSE132420, per mouse")
```

## Species without XIST

```{r species-status}
s <- SexSpecies()
s[
  s$scientific_name %in% c("Homo sapiens", "Mus musculus", "Rattus norvegicus", "Danio rerio"),
  c("scientific_name", "system", "inactivation_marker", "status")
]
```

### Rat: males only

Here we use [GSE100122](https://seqout.org/p/GSE100122), rat dorsal root ganglion, to show a
species whose panel has no `Xist`: males are called and females are `uncertain`.

```{r rat-bulk}
rb <- ExampleData("rat_bulk")
rb_cmp <- CompareSex(
  EstimateSex(rb$counts, species = rb$species, group = colnames(rb$counts)),
  rb$recorded
)
rb_cmp[, c("unit", "verdict", "confidence", "recorded", "flags")]
```

```{r rat-bulk-plot, fig.height = 4}
PlotSexEvidence(rb_cmp, main = "GSE100122, rat bulk")
```

### Zebrafish: no panel

Here we use [GSE123439](https://seqout.org/p/GSE123439), zebrafish gonads, to show a species
with no panel. This is strain AB, a lab strain that has lost the differentiated
sex chromosome, so every call is `unknown`.

```{r zebrafish}
zb <- ExampleData("zebrafish_bulk")
dim(zb$counts)
zb_cmp <- CompareSex(
  EstimateSex(zb$counts, species = zb$species, group = colnames(zb$counts)),
  zb$recorded
)
table(recorded = zb_cmp$recorded_sex, call = zb_cmp$verdict)
unique(zb_cmp$notes)
```

## Summary

```{r summary}
cmps <- list("human bulk" = hb_cmp, "mouse bulk" = mb_cmp)
if (has_tx) cmps[["mouse transcripts"]] <- mt_cmp
cmps <- c(cmps, list(
  "human single-cell" = hs_cmp, "mouse single-cell" = ms_cmp,
  "rat bulk" = rb_cmp, "zebrafish bulk" = zb_cmp
))
SummariseComparison(cmps)
```
